The Calculus of Linear Constructions

### 2021-12-02 writing/draft5/draft5.tex
draft5

% document layout
\documentclass[sigplan,screen,review,authordraft]{acmart}
%\documentclass[sigplan,screen]{acmart}
\usepackage[utf8]{inputenc}

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mdwtab}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{lstlangcoq}
\usepackage{float}
\usepackage{tikz-cd}
\usepackage{enumitem}

\lstset{basicstyle=\small,language=Coq,showstringspaces=false}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\setlist[itemize]{leftmargin=1.6em}
\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{\ $\text{let }#1\text{ := }#2\text{ in }#3$\ }
\newcommand{\ind}{\ $\text{ind}$\ }
\newcommand{\iter}{\ $\text{iter}$\ }
\newcommand{\unit}{\text{unit}}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\session}{\text{session}}
\newcommand{\channel}{\text{channel}}
\newcommand{\open}{\text{open}}
\newcommand{\close}{\text{close}}
\newcommand{\send}{\text{send}}
\newcommand{\recv}{\text{recv}}
\newcommand{\SEND}{\texttt{SEND}}
\newcommand{\RECV}{\texttt{RECV}}
\newcommand{\END}{\texttt{END}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}
\newcommand{\lift}[1]{\llparenthesis #1 \rrparenthesis}

% title and author
\title{The Calculus of Linear Constructions}
\author{Qiancheng Fu}
\affiliation{
  \institution{Boston University}
  \city{Boston}
  \state{MA}
  \country{USA}
}
\email{qcfu@bu.edu}

% document body
\begin{document}
\begin{abstract}
  The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends CC with a hierarchy of linear universes that precisely controls the weakening and contraction of its term level inhabitants. We study the meta-theory of CLC, showing that it is a sound logical framework for reasoning about resource. CLC is backwards compatible with CC, allowing CLC to enjoy the fruits of decades of CC research. We have formalized and proven correct all major results of the core calculus in the Coq Proof Assistant. We extend CLC with linear inductive types, showing that CLC facilitates correct by construction imperative programming.
\end{abstract}
\maketitle

\section{Introduction}
The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand and Huet in their landmark work \cite{cc}. In CC types can depend on terms, allowing one to write precise specifications as types. Today, CC and its variations CIC \cite{cic} and ECC \cite{ecc} lie at the core of popular proof assistants such as Coq \cite{coq}, Agda \cite{agda}, Lean \cite{lean} and others. These theorem provers have found great success in the fields of software verification, and constructive mathematics.

However, due to its origins as a logical framework for constructive mathematics, it is quite difficult for CC to encode and reason about resources. Intuitively, a mathematical theorem can be applied an unrestricted number of times. Comparatively, the usage of resources is more limited. For example, if we encode Girard's classical example \cite{girard95} of purchasing cigarettes literally into CC as a function of type:
\begin{equation*}
  Money \rightarrow Camels + Marlboro
\end{equation*}
If viewed as a propositional implication, the customer will still maintain full ownership of their money after paying the vendor, because implication does not diminish the validity of its antecedent. Unless the vendor is exceedingly generous, we are faced with the crime of counterfeiting. Users of proof assistants based on CC often need to embed external logics such as Separation Logic to provide additional reasoning principles for dealing with resource. The design and embedding of these logics are a difficult problem in their own right, requiring additional proofs to justify their soundness, and effectiveness. We propose an alternative solution: extend CC with linear types.

This paper presents a new linear dependent type system --- The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

@@

Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``lightweight'' approach to extending CC$\omega$ with linear types, akin to Mazurak et al.'s work on System F \cite{mazurak}. Prior works have either utilized a literal ! exponential or a pair of adjoint connectives to bridge the gap between linear and non-linear types, adding clutter that greatly hinders their practical usage. Our method requires neither exponential nor adjoint connectives, allowing for a straightforward modeling of CLC in CC$\omega$, and lifting of CC$\omega$ into CLC with minimal annotations, endowing CLC with the fruits of decades of CC research.

We further extend CLC with inductive types, showing that CLC facilitates correct by construction imperative programming. We have formalized all major results in Coq, and implemented a prototype in OCaml.

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, propagation and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
  \item Additionally, we observe that CLC is highly backwards compatible with CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We also prove that all valid CC$\omega$ terms after adding minimal amounts of annotation are valid CLC terms.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Furthermore, we extend CLC with linear inductive data, demonstrating CLC's capabilities for correct by construction imperative programming.
  \item Finally, we give an implementation extended with user definable linear and non-linear inductive types. Algorithmic type checking employed by the implementation streamlines the process of writing CLC.
\end{itemize}

\section{The Language of CLC}
\subsection{Syntax}
\begin{figure}[h]
  \vspace{-1em}
  \caption{Syntax}
  \vspace{1em}
  \begin{grammar}
    <$k$> := 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} predicative levels

    <$i$> := * | 0 | 1 | 2 ... \hspace*{3em} all levels

@@

    <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{4em} expressions
    \indalt $(x :_s A) \rightarrow B$
    \indalt $(x :_s A) \multimap B$
    \indalt $\lambda x. n$ | $m\ n$
  \end{grammar}
  \label{syntax}
\end{figure}
The syntax of the core type theory is presented in Figure \ref{syntax}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $k$ to specifically quantify over levels $0, 1, 2, ...$ that correspond to the predicative universes. We use the meta variable $i$ to quantify over all levels $*, 0, 1, 2, ...$. Here, $U_*$ is the impredicative universe of propositions, in the same spirit as CC$\omega$'s $Prop$ universe. $U_k$, and $L_k$ are the predicative universes of non-linear and linear types respectively.
\begin{figure}[h]
  \vspace{-1em}
  \caption{Correspondence of CLC types and MELL implications}
  \Description{}
  \vspace{-1em}
  \begin{align}
    (\_ \utype A) \rightarrow B \quad & \Leftrightarrow \quad !(!A \multimap B) \\
    (\_ \ltype A) \rightarrow B \quad & \Leftrightarrow \quad !(A \multimap B)  \\
    (\_ \utype A) \multimap B \quad   & \Leftrightarrow \quad !A \multimap B    \\
    (\_ \ltype A) \multimap B \quad   & \Leftrightarrow \quad A \multimap B
  \end{align}
  \vspace{-1em}
  \label{correspondence}
\end{figure}

A clear departure of our language from standard presentations of both linear type theory and dependent type theory is the presence of two function types: $(x :_s A) \rightarrow B$, $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic directly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theories. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we implicitly limit ! to only be canonically usable. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without requiring any additional machinery for manipulating exponential.

Figure \ref{correspondence} illustrates the correspondence between CLC function types and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (3) if the co-domain $B$ is dependent on arguments of domain $A$. We will discuss function type formation in Section \ref{tyformation} in greater detail.

In practice, algorithmic type checking techniques allow users to omit writing sort indices. Our implementation employs bi-directional type checking and users never interact with indices in the surface syntax.

\subsection{Universes and Cumulativity}
CLC features two sorts of universes $U$ and $L$ with level indices $*, 0, 1, 2, \cdots$. $U_*$ is the impredicative universe of propositions. $U_k$ and $L_k$ are the predicative universes of non-linear types and linear types respectively. The main mechanism that CLC uses to distinguish between linear and non-linear types is by checking the universe to which they belong. Basically, terms with types that occur within $U_i$ are unrestricted in their usage. Terms with types that occur within $L_k$ are restricted to being used exactly once.

@@

\begin{definition}
  The cumulativity relation ($\preceq$) is the smallest binary relation over terms such that
  \begin{enumerate}
    \item $\preceq$ is a partial order with respect to equality.
          \begin{enumerate}
            \item If $A \equiv B$, then $A \preceq B$.
            \item If $A \preceq B$ and $B \preceq A$, then $A \equiv B$.
            \item If $A \preceq B$ and $B \preceq C$, then $A \preceq B$.
          \end{enumerate}
    \item $U_* \preceq U_0 \preceq U_1 \preceq U_2 \preceq \cdots$
    \item $L_0 \preceq L_1 \preceq L_2 \preceq \cdots$
    \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, \\ then
          $(x \stype{s} A_1) \rightarrow B_1 \preceq (x \stype{s} A_2) \rightarrow B_2$
    \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, \\ then
          $(x \stype{s} A_1) \multimap B_1 \preceq (x \stype{s} A_2) \multimap B_2$
  \end{enumerate}
\end{definition}

Figure \ref{universe} illustrates the structure of our universe hierarchy. Each linear universe $L_k$ has $U_{k+1}$ as its type, allowing functions to freely quantify over linear \textit{types}. However, $L_k$ cumulates to $L_{k+1}$. These two parallel threads of cumulativity prevent linear types from being transported to the non-linear universes, and subsequently losing track of linearity.

\begin{figure}[H]
  \vspace{-1em}
  \caption{The Universe Hierarchy}
  \vspace{0.8em}
  \centering
  \begin{tikzcd}
    U_* \arrow[r, ":"', "\preceq", dash]
    & U_0 \arrow[r, ":"', "\preceq", dash]
    & U_1 \arrow[r, ":"', "\preceq", dash]
    & U_2 \arrow[r, ":"', "\preceq", dash]
    & \cdots \\
    & L_0 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash]
    & L_1 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash]
    & L_2 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash]
    & \cdots
  \end{tikzcd}
  \vspace{-0.5em}
  \label{universe}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \vspace{-1em}
  \caption{Structural Judgments}
  \vspace{-0.5em}
  \begin{mathpar}
    \inferrule
    { }
    { \epsilon \vdash }
    \rname{Wf-$\epsilon$}

@@

    \inferrule
    { \Gamma\ \vdash \\
      \overline{\Gamma} \vdash A \utype L_k }
    { \Gamma, x \ltype A\ \vdash }
    \rname{Wf-L}
    \\

@@

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A \utype U }
    { \pure{\Gamma, x \utype A} }
    \rname{Pure-U}
    \\

@@

    \inferrule
    { \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
      x \notin \Gamma_1 }
    { \mrg{\Gamma_1}
      {\Gamma_2, x \ltype A}
      {\Gamma, x \ltype A} }
    \rname{Merge-L2}
  \end{mathpar}
  \vspace{-1em}
  \label{structural}
  \Description{}
\end{figure}

@@

\begin{definition} We formally define the terms \textit{non-linear}, \textit{linear}, \textit{unrestricted} and \textit{restricted}.
  \begin{enumerate}
    \item A \textit{type} $A$ is \textit{non-linear} under context $\Gamma$ if $\Gamma \vdash A : U_i$.
    \item A \textit{type} $A$ is \textit{linear} under context $\Gamma$ if $\Gamma \vdash A : L_k$.
    \item A \textit{term} $m$ is \textit{unrestricted} under context $\Gamma$ if there exists non-linear type $A$, such that $\Gamma \vdash m : A$. A unrestricted term may be used an arbitrary number of times.
    \item A \textit{term} $m$ is \textit{restricted} under context $\Gamma$ if there exists linear type $A$, such that $\Gamma \vdash m : A$. A restricted term must be used exactly once.
  \end{enumerate}
\end{definition}

@@

The axiom rules \rname{Prop-Axiom}, \rname{U-Axiom}, \rname{L-Axiom} are almost standard, the main difference being the extra side-condition of judgment $\pure{\Gamma}$. In most presentations of dependent type theories without linear types, the universe axioms are derivable under any well-formed context $\Gamma$. Variables not pertaining to actual proofs could be introduced this way, thus giving rise to the admissibility of weakening. To support linear types, we must restrict weakening to non-linear variables. This justifies the restriction of $\Gamma$ to contain only non-linear variables for \rname{Prop-Axiom}, \rname{U-Axiom}, \rname{L-Axiom}. From \rname{L-Axiom} we can see that the universe of linear types $L_k$ is an inhabitant of $U_{k+1}$. This is inspired by Krishnaswami et al.'s treatment of linear universes \cite{neel15}, where linear \textit{types} themselves can be used unrestrictedly.

The \rname{Prop} rule is used for forming propositions. From the judgment $\Gamma \vdash A : U_i$ we can see that \rname{Prop} allows impredicative quantification over non-linear types of arbitrary level $i$. The judgment $\Gamma, x \utype A \vdash B : U_*$ asserts that the co-domain must be in the impredicative universe $U_*$. The final resulting judgment $\Gamma \vdash (x \utype A) \rightarrow B : U_*$ has sort $U$, indicating that terms with type $(x \utype A) \rightarrow B$ enjoy unrestricted usage.

\begin{figure}[h]
  \caption{Type Formation}
  \begin{mathpar}
    \inferrule
    { \pure{\Gamma} }
    { \Gamma \vdash U_* : U_0 }
    \rname{Prop-Axiom}

    \inferrule
    { \pure{\Gamma} }
    { \Gamma \vdash s_k : U_{k+1} }
    \rname{Sort-Axiom}

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : U_i \\
      \Gamma, x \utype A \vdash B : U_* }
    { \Gamma \vdash (x \utype A) \rightarrow B : U_* }
    \rname{Prop}
    \\

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : U_k \\
      \Gamma, x \utype A \vdash B : s_k }
    { \Gamma \vdash (x \utype A) \rightarrow B : U_k }
    \rname{U-Prod}

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : L_k \\
      \Gamma \vdash B : s_k \\
      x \notin \Gamma }
    { \Gamma \vdash (x \ltype A) \rightarrow B : U_k }
    \rname{L-Prod}
    \\

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : U_k \\
      \Gamma, x \utype A \vdash B : s_k }
    { \Gamma \vdash (x \utype A) \multimap B : L_k }
    \rname{U-Lolli}

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : L_k \\
      \Gamma \vdash B : s_k \\
      x \notin \Gamma }
    { \Gamma \vdash (x \ltype A) \multimap B : L_k }
    \rname{L-Lolli}
  \end{mathpar}
  \label{type}
  \Description{}
\end{figure}

>>>>

\documentclass[sigplan,screen,review,authordraft]{acmart}

\usepackage[utf8]{inputenc}

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mdwtab}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{float}
\usepackage{tikz-cd}
\usepackage{enumitem}

\setlist[itemize]{leftmargin=1.6em}
\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{\ $\text{let }#1\text{ := }#2\text{ in }#3$\ }
\newcommand{\ind}[1]{\text{Ind}_{#1}}
\newcommand{\constr}{\text{Constr}}
\newcommand{\case}{\text{Case}}
\newcommand{\dcase}{\text{DCase}}
\newcommand{\fix}{\text{Fix }}
\newcommand{\sep}{\text{ | }}
\newcommand{\unit}{\text{unit}}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\session}{\text{session}}
\newcommand{\channel}{\text{channel}}
\newcommand{\open}{\text{open}}
\newcommand{\close}{\text{close}}
\newcommand{\send}{\text{send}}
\newcommand{\recv}{\text{recv}}
\newcommand{\SEND}{\texttt{SEND}}
\newcommand{\RECV}{\texttt{RECV}}
\newcommand{\END}{\texttt{END}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}
\newcommand{\lift}[1]{\llparenthesis #1 \rrparenthesis}
\newcommand{\lrangle}[1]{\langle #1 \rangle}
\newcommand{\ucons}{constructor_{\scriptscriptstyle U}}
\newcommand{\lcons}{constructor_{\scriptscriptstyle L}}
\newcommand{\scons}{constructor_{s}}
\newcommand{\inl}{\text{inl}}
\newcommand{\inr}{\text{inr}}

\title{The Calculus of Linear Constructions}
\author{Qiancheng Fu}
\affiliation{
  \institution{Boston University}
  \city{Boston}
  \state{MA}
  \country{USA}
}
\email{qcfu@bu.edu}

\keywords{Dependent types, linear types, inductive types, Calculus of Constructions}

\begin{document}
\begin{abstract}
  The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends the predicative CC$\omega$ with a hierarchy of linear universes that precisely controls the weakening and contraction of its term level inhabitants. We study the meta-theory of CLC, showing that it is a sound logical framework for reasoning about resource. We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Through examples, we demonstrate that CILC facilitates correct by construction imperative programming and lightweight protocol enforcement. We have formalized and proven correct all major results in the Coq Proof Assistant.
\end{abstract}
\maketitle

\section{Introduction}
The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand and Huet in their landmark work \cite{cc}. In CC types can depend on terms, allowing one to write precise specifications as types. Today, CC and its extensions CIC \cite{cic} and ECC \cite{ecc} lie at the core of popular proof assistants such as Coq \cite{coq}, Agda \cite{agda}, Lean \cite{lean} and others. These theorem provers have found great success in the fields of software verification, and constructive mathematics.

However, due to its origins as a logical framework for constructive mathematics, it is quite difficult for CC to encode and reason about resources. Intuitively, a mathematical theorem can be applied an unrestricted number of times. Comparatively, the usage of resources is more limited. For example, if we encode Girard's classical example \cite{girard95} of purchasing cigarettes literally into CC as a function of type:
\begin{equation*}
  Money \rightarrow Camels + Marlboro
\end{equation*}
If viewed as a propositional implication, the customer will still maintain full ownership of their money after paying the vendor, because implication does not diminish the validity of its antecedent. Unless the vendor is exceedingly generous, we are faced with the crime of counterfeiting. Users of proof assistants based on CC often need to embed external logics such as Separation Logic to provide domain-specific reasoning principles for dealing with resource. We propose an alternative solution: extend CC with linear types.

This paper presents a new linear dependent type system --- The Calculus of Linear Constructions (CLC). CLC extends the predicative CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

@@

Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``lightweight'' approach, akin to Mazurak et al.'s work on System F \cite{mazurak}. Prior works have either utilized an explicit ! exponential or a pair of adjoint connectives to mediate between the worlds of linear and non-linear types. Our method requires neither exponential nor adjoint connectives explicitly in the syntax, allowing for a straightforward modeling of CLC in the predicative CC$\omega$, and lifting of predicative CC$\omega$ into CLC with minimal annotations needed.

We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Linear connectives such a $\otimes$ and $\oplus$ are definable using this generalized mechanism in a straightforward way. Through examples, we show that CILC facilitates correct by construction imperative programming and lightweight protocol enforcement.

We have formalized all major results in Coq, and implemented a prototype in OCaml, both of which are available in the first author's Github repository.

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, validity and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
  \item Additionally, we observe that CLC is backwards compatible with the predicative CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We prove that all valid CC$\omega$ terms are valid CLC terms after adding minimal amounts of annotation.
  \item Furthermore, we extend CLC with rules to define inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). This is the first formalization of a generalized mechanism for defining inductive linear types that we are aware of.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Finally, we give an implementation of CILC in OCaml. Algorithmic type checking employed by the typechecker streamlines the process of writing CILC programs.
\end{itemize}

\section{The Language of CLC}
\subsection{Syntax}
\begin{figure}[h]
  \caption{Syntax of CLC}
  \begin{grammar}
    <$i$> := 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} universe levels

@@

    <$m, n, A, B, M$> ::= $U_i$ | $L_i$ | $x$ \hspace*{4.2em} expressions
    \indalt $(x :_s A) \rightarrow B$
    \indalt $(x :_s A) \multimap B$
    \indalt $\lambda x :_s A. n$
    \indalt $m\ n$
  \end{grammar}
  \Description{}
  \label{syntax1}
\end{figure}

The syntax of the core type theory is presented in Figure \ref{syntax1}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $i$ to quantify over universe levels $0, 1, 2, ...$. $U_i$, and $L_i$ are the universes at level $i$ of non-linear and linear types respectively.
\begin{figure}[h]
  \caption{Correspondence of CLC types and MELL implications}
  \Description{}
  \begin{align}
    (\_ \utype A) \rightarrow B \quad & \Leftrightarrow \quad !(!A \multimap B) \\
    (\_ \ltype A) \rightarrow B \quad & \Leftrightarrow \quad !(A \multimap B)  \\
    (\_ \utype A) \multimap B \quad   & \Leftrightarrow \quad !A \multimap B    \\
    (\_ \ltype A) \multimap B \quad   & \Leftrightarrow \quad A \multimap B
  \end{align}
  \label{correspondence}
\end{figure}

A clear departure of our language from standard presentations of both linear type theory and dependent type theory is the presence of two function types: $(x :_s A) \rightarrow B$, $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic implicitly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theories. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we have limited ! to only be canonically usable. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without requiring any additional machinery for manipulating exponential.

Figure \ref{correspondence} illustrates the correspondence between CLC functions and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (3) if the co-domain $B$ is dependent on arguments of domain $A$. We will discuss function type formation in Section \ref{tyformation} in greater detail.

Algorithmic type checking techniques allow users to omit writing sort indices for many situations in practice. Our implementation employs bi-directional type checking and users rarely interact with sort indices in the surface syntax.

\subsection{Universes and Cumulativity}
CLC features two sorts of universes $U$ and $L$ with level indices $0, 1, 2, \cdots$. $U_i$ and $L_i$ are the predicative universes of non-linear types and linear types respectively. The main mechanism that CLC uses to distinguish between linear and non-linear types is by checking the universe to which they belong. Basically, terms with types that occur within $U_i$ are unrestricted in their usage. Terms with types that occur within $L_i$ are restricted to being used exactly once.

@@

\begin{definition}
  The cumulativity relation ($\preceq$) is the smallest binary relation over terms such that
  \begin{enumerate}
    \item $\preceq$ is a partial order with respect to equality.
          \begin{enumerate}
            \item If $A \equiv B$, then $A \preceq B$.
            \item If $A \preceq B$ and $B \preceq A$, then $A \equiv B$.
            \item If $A \preceq B$ and $B \preceq C$, then $A \preceq B$.
          \end{enumerate}
    \item $U_0 \preceq U_1 \preceq U_2 \preceq \cdots$
    \item $L_0 \preceq L_1 \preceq L_2 \preceq \cdots$
    \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, \\ then
          $(x \stype{s} A_1) \rightarrow B_1 \preceq (x \stype{s} A_2) \rightarrow B_2$
    \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, \\ then
          $(x \stype{s} A_1) \multimap B_1 \preceq (x \stype{s} A_2) \multimap B_2$
  \end{enumerate}
\end{definition}

Figure \ref{universe} illustrates the structure of our universe hierarchy. Each linear universe $L_i$ has $U_{i+1}$ as its type, allowing functions to freely quantify over linear \textit{types}. However, $L_i$ cumulates to $L_{i+1}$. These two parallel threads of cumulativity prevent linear types from being transported to the non-linear universes, and subsequently losing track of linearity.

\begin{figure}[h]
  \caption{The Universe Hierarchy}
  \vspace{1em}
  \centering
  \begin{tikzcd}
    U_0 \arrow[r, ":"', "\preceq", dash]
    & U_1 \arrow[r, ":"', "\preceq", dash]
    & U_2 \arrow[r, ":"', "\preceq", dash]
    & \cdots \\
    L_0 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash]
    & L_1 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash]
    & L_2 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash]
    & \cdots
  \end{tikzcd}
  \label{universe}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \caption{Structural Judgments}
  \begin{mathpar}
    \inferrule
    { }
    { \epsilon \vdash }
    \rname{Wf-$\epsilon$}

@@

    \inferrule
    { \Gamma\ \vdash \\
      \overline{\Gamma} \vdash A \utype L_k }
    { \Gamma, x \ltype A\ \vdash }
    \rname{Wf-L}

@@

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A \utype U }
    { \pure{\Gamma, x \utype A} }
    \rname{Pure-U}

@@

    \inferrule
    { \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
      x \notin \Gamma_1 }
    { \mrg{\Gamma_1}
      {\Gamma_2, x \ltype A}
      {\Gamma, x \ltype A} }
    \rname{Merge-L2}
  \end{mathpar}
  \label{structural}
  \Description{}
\end{figure}

@@

\begin{definition} We formally define the terminology \textit{non-linear}, \textit{linear}, \textit{unrestricted} and \textit{restricted}.
  \begin{enumerate}
    \item A \textit{type} $A$ is \textit{non-linear} under context $\Gamma$ if $\Gamma \vdash A : U_i$.
    \item A \textit{type} $A$ is \textit{linear} under context $\Gamma$ if $\Gamma \vdash A : L_k$.
    \item A \textit{term} $m$ is \textit{unrestricted} under context $\Gamma$ if there exists non-linear type $A$, such that $\Gamma \vdash m : A$. A unrestricted term may be used an arbitrary number of times.
    \item A \textit{term} $m$ is \textit{restricted} under context $\Gamma$ if there exists linear type $A$, such that $\Gamma \vdash m : A$. A restricted term must be used exactly once.
  \end{enumerate}
\end{definition}

@@

The axiom rules \rname{U-Axiom}, \rname{L-Axiom} are almost standard, the main difference being the extra side-condition of judgment $\pure{\Gamma}$. In most presentations of dependent type theories without linear types, the universe axioms are derivable under any well-formed context $\Gamma$. Variables not pertaining to actual proofs could be introduced this way, thus giving rise to the admissibility of weakening. To support linear types, we must restrict weakening to non-linear variables. This justifies the restriction of $\Gamma$ to contain only non-linear variables for \rname{U-Axiom}, \rname{L-Axiom}. From \rname{L-Axiom} we can see that the universe of linear types $L_i$ is an inhabitant of $U_{i+1}$. This is inspired by Krishnaswami et al.'s treatment of linear universes \cite{neel15}, where linear \textit{types} themselves can be used unrestrictedly.

\begin{figure}[h]
  \caption{Type Formation}
  \begin{mathpar}
    \inferrule
    { \pure{\Gamma} }
    { \Gamma \vdash s_i : U_{i+1} }
    \rname{Sort-Axiom}

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : U_i \\
      \Gamma, x \utype A \vdash B : s_i }
    { \Gamma \vdash (x \utype A) \rightarrow B : U_i }
    \rname{U$\rightarrow$}

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : L_i \\
      \Gamma \vdash B : s_i \\
      x \notin \Gamma }
    { \Gamma \vdash (x \ltype A) \rightarrow B : U_i }
    \rname{L$\rightarrow$}

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : U_i \\
      \Gamma, x \utype A \vdash B : s_i }
    { \Gamma \vdash (x \utype A) \multimap B : L_i }
    \rname{U$\multimap$}

    \inferrule
    { \pure{\Gamma} \\
      \Gamma \vdash A : L_i \\
      \Gamma \vdash B : s_i \\
      x \notin \Gamma }
    { \Gamma \vdash (x \ltype A) \multimap B : L_i }
    \rname{L$\multimap$}
  \end{mathpar}
  \label{type}
  \Description{}
\end{figure}

The \rname{U$\rightarrow$} rule is used for forming non-linear function types with non-linear domains. This is evident from the judgment $\Gamma \vdash A : U_i$. Due to the fact that $A$ is in the non-linear universe $U_i$, terms of type $A$ have unrestricted usage. The non-linearity of domain $A$ allows $B$ to depend on terms of type $A$, as seen in judgment $\Gamma, x \utype A \vdash B : s_i$. The domain $B$ itself may be non-linear or linear, since $B$'s universe $s_k$ can vary between $U_i$ and $L_i$. From the resulting judgment $\Gamma \vdash (x \utype A) \rightarrow B : U_i$, we see that the overall type is a non-linear type. The term level $\lambda$-abstractions of these types can be used unrestrictedly.
